Competing Bandits: The Perils of Exploration under Competition

### 2018-09-03 paper/content/barriers.tex
updates

\swedit{To answer these questions, we isolate the effects from the two
  advantages and run two additional experiments. In the first
  experiment (called \emph{reputation erased experiment}), the
  reputation of the incumbent is artificially reset upon the entry of
  the entrant \gaedit{so that the incumbent's reputation score is only set by the warm start periods and not any of the incumbency periods}. In the second experiment (called \emph{information
    erased experiment}), the information gained by the incumbent is
  erased in the sense that its posterior is reset to the initial
  (fake) prior upon the entry of the entrant.}

@@

\begin{table}[ht]
\centering
\caption{Information Erased Experiment Heavy Tail} 
\begin{tabular}{c@{\hspace{0.1\tabcolsep}}ccc}
\hline
\multicolumn{4}{c}{Incumbent Algorithm}\\
\multirow{12}{0.6in}{\rotatebox{90}{Entrant Algorithm}} \\
  \hline
 & TS & DEG &  DG \\ 
  \hline
TS & \makecell{\textbf{0.021} $\pm$0.009\\Var:0.02\\ES:100\%} & \makecell{\textbf{0.16} $\pm$0.02\\Var:0.1\\ES:97\%} & \makecell{\textbf{0.21} $\pm$0.02\\Var:0.2\\ES:95\%} \\ 
  DEG & \makecell{\textbf{0.26} $\pm$0.03\\Var:0.2\\ES:95\%} & \makecell{\textbf{0.3} $\pm$0.02\\Var:0.2\\ES:74\%} & \makecell{\textbf{0.26} $\pm$0.02\\Var:0.1\\ES:76\%} \\ 
   DG & \makecell{\textbf{0.34} $\pm$0.03\\Var:0.2\\ES:94\%} & \makecell{\textbf{0.4} $\pm$0.03\\Var:0.2\\ES:74\%} & \makecell{\textbf{0.33} $\pm$0.02\\Var:0.1\\ES:58\%} \\ 
   \hline
  \label{info_erase}
\end{tabular}
\end{table}

\begin{table}[ht]
\centering
\caption{Reputation Erased Experiment Heavy Tail} 
\begin{tabular}{c@{\hspace{0.01\tabcolsep}}ccc}
\hline
\multicolumn{4}{c}{Incumbent Algorithm}\\
\multirow{12}{0.6in}{\rotatebox{90}{Entrant Algorithm}} \\
  \hline
 & TS & DEG &  DG \\ 
  \hline
TS & \makecell{\textbf{0.0096} $\pm$0.006\\Var:0.009\\ES:100\%} & \makecell{\textbf{0.11} $\pm$0.02\\Var:0.09\\ES:98\%} & \makecell{\textbf{0.18} $\pm$0.02\\Var:0.1\\ES:95\%} \\ 
  DEG & \makecell{\textbf{0.073} $\pm$0.01\\Var:0.05\\ES:93\%} & \makecell{\textbf{0.29} $\pm$0.02\\Var:0.2\\ES:78\%} & \makecell{\textbf{0.25} $\pm$0.02\\Var:0.1\\ES:79\%} \\ 
   DG & \makecell{\textbf{0.15} $\pm$0.02\\Var:0.1\\ES:89\%} & \makecell{\textbf{0.39} $\pm$0.03\\Var:0.2\\ES:78\%} & \makecell{\textbf{0.33} $\pm$0.02\\Var:0.2\\ES:66\%} \\ 
   \hline
    \label{rep_erase}
\end{tabular}
\end{table}

\begin{finding}
\textit{Maintaining either the data or reputation advantage alone still allows the incumbent \gaedit{to retain a large share of the market compared to the share they get under permanent duopoly}. \swcomment{too vague. what does significant mean?}\\ \indent When the incumbent plays $\TS$, the data advantage serves as a larger barrier to entry compared to reputation, but when the incumbent plays $\DEG$ or $\DG$ there is no substantial difference between the two barriers.}
\end{finding}

@@

\gadelete{Though the setup is purely experimental, it is nonetheless interesting to look at if the same strategies remain best responses in this setting compared to the setting where both data and reputation are retained. We find that the best responses remain the same for the Heavy Tail instance, but for the Uniform instance they change where $\TS$ is weakly dominant for the incumbent under the reputation erased treatment but not under the information erased treatment.}

>>>>

\swedit{To answer these questions, we isolate the effects from the two
  advantages and run two additional experiments. In the first
  experiment (called \emph{reputation erased experiment}), the
  reputation of the incumbent is artificially reset upon the entry of
  the entrant \gaedit{so that the incumbent's reputation score is not influenced by the rewards during the incumbency periods and only determined by the rewards generated in the warm start periods}. In the second experiment (called \emph{information
    erased experiment}), the information gained by the incumbent is
  erased in the sense that its posterior is reset to the initial
  (fake) prior upon the entry of the entrant.}

@@

\begin{table}[ht]
\centering
\caption{Information Erased Experiment Heavy Tail}
\begin{tabular}{cc|c|c|c|}
  & \multicolumn{1}{c}{} & \multicolumn{3}{c}{Incumbent} \\
  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\TS$}  & \multicolumn{1}{c}{$\DEG$}  & \multicolumn{1}{c}{$\DG$} \\\cline{3-5}
            & $\TS$ & \makecell{\textbf{0.021} \\$\pm$0.009}& 		\makecell{\textbf{0.16} \\$\pm$0.02} & 
            \makecell{\textbf{0.21} \\ $\pm$0.02} \\ \cline{3-5}
Entrant  & $\DEG$ & \makecell{\textbf{0.26} \\$\pm$0.03} & \makecell{\textbf{0.3} \\$\pm$0.02} & 
\makecell{\textbf{0.26} \\ $\pm$0.02} \\ \cline{3-5}
            & $\DG$ & \makecell{\textbf{0.34} \\ $\pm$0.03} & \makecell{\textbf{0.4} \\$\pm$0.03} & 
            \makecell{\textbf{0.33} \\$\pm$0.02} \\\cline{3-5}
\end{tabular}
\label{info_erase}
\end{table}

\begin{table}[ht]
\centering
\caption{Reputation Erased Experiment Heavy Tail}
\begin{tabular}{cc|c|c|c|}
  & \multicolumn{1}{c}{} & \multicolumn{3}{c}{Incumbent} \\
  & \multicolumn{1}{c}{} & \multicolumn{1}{c}{$\TS$}  & \multicolumn{1}{c}{$\DEG$}  & \multicolumn{1}{c}{$\DG$} \\\cline{3-5}
            & $\TS$ & \makecell{\textbf{0.0096} \\$\pm$0.006}& 		\makecell{\textbf{0.11} \\$\pm$0.02} & 
            \makecell{\textbf{0.18} \\$\pm$0.02} \\ \cline{3-5}
Entrant  & $\DEG$ & \makecell{\textbf{0.073} \\ $\pm$0.01} & \makecell{\textbf{0.29} \\ $\pm$0.02} & 
\makecell{\textbf{0.25} \\$\pm$0.02} \\ \cline{3-5}
            & $\DG$ &\makecell{\textbf{0.15} \\$\pm$0.02} & \makecell{\textbf{0.39} \\$\pm$0.03} & 
           \makecell{\textbf{0.33} \\ $\pm$0.02} \\\cline{3-5}
\end{tabular}
\label{rep_erase}
\end{table}

\begin{finding}
\textit{Maintaining either the data or reputation advantage alone still allows the incumbent \gaedit{to retain a large share of the market compared to the share they get under permanent duopoly}.\\ \indent When the incumbent plays $\TS$, the data advantage serves as a larger barrier to entry compared to reputation, but when the incumbent plays $\DEG$ or $\DG$ there is no substantial difference between the two barriers.}
\end{finding}

@@

\OMIT{Though the setup is purely experimental, it is nonetheless interesting to look at if the same strategies remain best responses in this setting compared to the setting where both data and reputation are retained. We find that the best responses remain the same for the Heavy Tail instance, but for the Uniform instance they change where $\TS$ is weakly dominant for the incumbent under the reputation erased treatment but not under the information erased treatment.}

### 2018-09-03 paper/content/revisited.tex
updates

Conjecture~\ref{conj:inverted-u} for the most part is validated by our simulation results, but Puzzles~\ref{puz:dg_ht} and \ref{puz:nih} seem to be problematic for it.

We first attempt to explain Puzzle~\ref{puz:dg_ht} by revisiting Conjecture~\ref{conj:mean-trajectories}.

\begin{finding}
\textit{
Conjecture~\ref{conj:mean-trajectories} is false. Mean reputation trajectories do not suffice to explain the outcomes under competition.}
\end{finding}

@@

Let us consider a statistic with a more direct connection to the competition game: for each time $t$, the proportion of \MRVs for which one algorithm has a higher reputation score than the other.

\begin{definition}
\textit{Relative Reputation Proportion} - the proportion of simulations in which algorithm $A$ had at least as high of a reputation as algorithm $B$ for a fixed time $t$
\end{definition}

\begin{finding}
\textit{Purposeful exploration can lead to relative reputational costs compared to the greedy alternative and this leads to $\TS$ doing worse than $\DG$ for small time horizons. We observe this under the Uniform and Heavy Tail instances. However, it is not observed under the Needle In Haystack instances}
\end{finding}

The relative reputation proportion statistic corresponds to running the bandit algorithms in isolation on the same instance and with the same realizations for $t$ rounds and then calculating the fraction of simulations at which an agent would select a firm playing $A$ over a firm playing $B$ at time $t$.

\begin{figure}[ht]
\caption{Relative Reputation Proportion Plots for $\TS$ vs $\DG$. Shaded area display 95\% confidence intervals.}
\includegraphics[scale=0.35]{figures/relative_uniform_annotated_plot}
\includegraphics[scale=0.35]{figures/relative_nih_ts_dg.png}
%\caption*{\tiny{The plots contain the average reputation over $1000$ runs for a memory size of $100$ where, for a given $t$, we record the reputation of both of the algorithms on a given instance and then calculate the proportion of runs where $\TS \geq \DG$. The shaded area display 95\% confidence intervals.}}
\label{relative_rep_plots}

\end{figure}

Figure \ref{relative_rep_plots} shows plots of the relative reputation proportion for $\TS$ vs $\DG$ on the Uniform and Needle In Haystack instances. For the Uniform instance we see that, in the early rounds, the relative reputation proportion for $\TS$  vs. $\DG$ is less than 0.5 meaning that $\DG$ has a higher reputation in a majority of the simulations  but that, eventually, the proportion $\TS$ does better than $\DG$ in a majority of the simulations. The intuition behind this is that, especially since the firms start with no substantive initial information, $\TS$ does purposeful exploration in the early rounds in order to acquire information. However, eventually the information acquired from purposeful exploration in the early rounds allows $\TS$ to make better decisions and achieve a higher reputation, especially when the instance is ``hard enough" so that $\DG$ cannot trivially find the best arm. The early exploration leads to what we define as a \textit{relative reputation cost} and the eventual gain in reputation leads to what we define as a \textit{relative reputation benefit}.

\begin{definition}
\textit{Relative reputation cost and benefit} - the relative reputation loss an algorithm incurs from purposeful exploration compared to the greedy alternative. Here we treat it as an empirical definition based on the relative reputation proportion where the ``cost" regime is when the proportion $< 0.5$ and the ``benefit" regime is when the proportion is $> 0.5$ for the better algorithm. We call the priors where there is an early costly period followed by a later benefit period relative reputation costly instances.
\end{definition}

Is exploration always costly? Figure \ref{relative_rep_plots} also shows that for the Needle In Haystack instances, $\TS$ always does relatively better than $\DG$. There are two contributing factors to this. First, $\TS$ identifies the best arm faster in the Needle In Haystack prior than the Uniform prior so that there is a shorter time horizon where $\TS$ needs to engage in purposeful exploration. Second, in the Needle In Haystack prior there are no ``bad" arms as there may be in the Uniform prior since by construction all the arms except one in Needle In Haystack are the same. Thus, when $\TS$ pulls a sub-optimal arm relative to its current information, the expected reward is the same as the greedy option that has not identified the best arm. However, with the Uniform prior, it is possible that the sub-optimal arm that is pulled has substantially lower expected reward relative to the greedy option. Thus, only the Uniform and Heavy Tail instances are relative reputation costly priors.

This observation allows us to explain the Puzzle~\ref{puz:nih} from the competition game and establish a more nuanced view of the inverted-U.

\begin{finding}
\textit{For sufficiently low warm start \emph{under the instances that are relative reputation costly}, the optimal strategies in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $DG$ \\
\textbf{Temporary Monopoly} - $\TS$ \\
\textbf{Permanent Duopoly} - $DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{finding}

Thus, the reason we see that $\DG > \TS$ under Heavy Tail and Uniform but that $\TS$ is always preferred under Needle In Haystack is due to the fact that the purposeful exploration engaged by $\TS$ in the early rounds leads it to suffer a relative reputational cost in the former case but not in the latter. Looking at the relative reputation plots in Figure \ref{relative_rep_plots}, we can interpret fixing a warm start $T_0$ as fixing the starting point on the relative reputation plots. The proportion of first rounds in the competition game that will go to a firm playing alg $A$ over a firm playing alg $B$ will correspond to the relative reputation proportion at time $T_0$. As a result, if we begin with a warm start that falls in the region where the proportion $< 0.5$, we expect that $\DG > \TS$.

This also allows us to give some intuition as to why allowing one firm to be a temporary monopoly incentivizes it to play $\TS$ even if, for the same warm start, the firm would be incentivized to play $\DG$ under permanent duopoly.  By allowing it to be the only firm in the market for sufficiently long we allow it to recoup the ``reputation costs" by having it engage in most of its exploration while it does not have the pressure to incentivize agents to select them over their competition.

>>>>

A natural question to ask is why we had to introduce the relative reputation proportion instead of simply using the mean to understand the results. For instance, the following would have been a natural conjecture:
\begin{conjecture}\label{conj:mean-trajectories}
\begin{enumerate}
\item To understand which algorithm would win in the competition game we could compare the mean reputation score of the two algorithms at $T_0$.
\item Given a sufficiently large $T_0$, if one algorithm's mean reputation trajectory lies above another, then the first algorithm prevails in the competition game.
\end{enumerate}
\end{conjecture}

However, we find a more nuanced picture.

\begin{finding}
\textit{
Conjecture~\ref{conj:mean-trajectories} is false. Mean reputation trajectories do not suffice to explain the outcomes under competition. We observe in Figure \ref{sim_ht} that $DG$ attains a larger market share than $DEG$ even for large warm starts.\footnote{In the appendix we show that this holds for $K = 3$ and longer time horizons than we consider here as well}}
\end{finding}

@@
